### Analyzing ACS microdata

What was the gender wage-gap in Oakland in

```{r}
#| code-fold: true
individuals_w_income <- oak %>%
  # Find adult earners
  filter(INCTOT != 9999999, INCTOT > 0, AGE >= 18) %>%
  mutate(
    # Label sex
    SEX = case_when(SEX == 1 ~ 'Male', TRUE ~ 'Female'),
    # Label education
    educ_attain = case_when(
      EDUC == 10 ~ "Bachelor's degree",
      EDUC == 11 ~ "Graduate degree",
      EDUCD %in% c(63, 65, 64) ~ "Highschool diploma",
      EDUCD == 71 ~ "Some college",
      EDUC == 8 ~ "Associate's degree",
      EDUC == 0 ~ "No schooling",
      EDUCD == 61 ~ "Some school",
      EDUC < 6 ~ "Some school",
    ) %>% as.factor()
  )

gof_stuff <- tribble(
  ~ raw, ~ clean, ~ fmt,
  "nobs", "N", 0,
  "r.squared", "R²", 3
  )
```

```{r}
oak_indiv_svy <- as_survey_rep(
  individuals_w_income,
  weight = PERWT ,
  repweights = matches("REPWTP[0-9]+"),
  type = "ACS",
  mse = TRUE)
```

```{r}
oak_indiv_svy %>%
  group_by(SEX) %>%
  summarize(mean_income = survey_mean(INCTOT))
```

```{r}
#| warning: false
#| output: false
#| echo: false
library(reticulate)
use_condaenv('base')
```

```{python}
#| output: false
#| echo: false
#| 
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns

df = pd.DataFrame(r['individuals_w_income'])

fig, ax = plt.subplots(figsize=(3, 3))
sns.kdeplot(data=df, x='INCTOT', y="AGE", weights="PERWT", ax=ax, fill=True, alpha=.5, log_scale=True, legend=False, cmap="coolwarm")
ax.set_axis_off()
ax.set_xlim(10e2, 10e5)
fig.tight_layout()
fig.savefig("thumbnail.png", dpi=300)
plt.show();
```

Lets explore a couple other dynamics related to earnings in Oakland and
practice making visualizations and estimating regressions with the data.
Here's the income distribution across different levels of highest
education earned. Note here that I'm not using the replicate weights
(I'm just using the `individuals_w_income` object), since I'm not
estimating anything. `ggplot` does take a weight argument, for which
I've supplied the `PERWT` sample weights.

<https://www.andrewheiss.com/blog/2022/06/23/long-labels-ggplot/index.html>

```{r}
#| warning: false
#| code-fold: true

order <- individuals_w_income %>%
  distinct(EDUC, educ_attain) %>%
  arrange(EDUC) %>%
  distinct(educ_attain) %>%
  pull()

ggplot(individuals_w_income, aes(
  x = factor(educ_attain, levels = order),
  y = INCTOT,
  weight = PERWT
)) +
  geom_jitter(
    position = position_jitter(width = .2),
    alpha = 0.6,
    color = "grey",
    size = 1.2
  ) +
  geom_boxplot(
    alpha = 0.9,
    color = "black",
    size = .9,
    outliers = FALSE,
    linewidth = .8
  ) +
  scale_y_continuous(
    labels = scales::label_currency(scale_cut = scales::cut_short_scale()),
    limits = c(0, 500000),
    breaks = seq(0, 500000, 100000)
  ) +
  scale_x_discrete(labels = scales::label_wrap(10)) +
  labs(title = "Income by Education Level", y = "Income", x = "Education Level") +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid = element_line(
      color = "lightgrey",
      size = .2,
      linetype = 1
    ),
    panel.background = element_rect("white")
  )
```

The plot shows that individuals with bachelor's and graduate degrees
generally have higher incomes in Oakland. Let's seen how bachelor and
graduate degree attainment differs across sex. Again, I'm using the
`individuals_w_income` object as I have no need for replicate weights.

```{r}
#| warning: false
#| code-fold: true

colors <- RColorBrewer::brewer.pal(n = 5, "Set1")[c(5, 2)]

order <- individuals_w_income %>% distinct(EDUC, educ_attain) %>% arrange(desc(EDUC)) %>% distinct(educ_attain) %>% pull()

ggplot(individuals_w_income,
       aes(
         y = factor(educ_attain, level = order),
         color = factor(SEX),
         fill = factor(SEX),
         weight = PERWT
       )) +
  geom_bar(
    position = "dodge",
    boundary = 0,
    alpha = 0.9,
    aes(x = (..count..) / sum(..count..))
  ) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_x_continuous(labels = scales::label_percent(), ) +
  labs(
    title = "Distribution of Highest Education Level by Sex",
    x = "Percent of earners",
    y = "Education",
    color = "Sex",
    fill = "Sex"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid = element_line(
      color = "lightgrey",
      size = .2,
      linetype = 1
    ),
    panel.background = element_rect("white")
  )
```

So bachelor and graduate degree holders generally earn more, and women
are more likely than men to have bachelor and graduate degrees in
Oakland. This should imply that if we adjust for education when
estimating the gender-earnings gap in Oakland, thus only comparing men
and women in the same education levels, we should see the gap increase.

We again use `svyglm` to estimate the regression, and `oak_indiv_svy` as
the data/survey design so as to use the replicate weights for our
standard errors.

```{r}
model_ols2 <- svyglm(log(INCTOT) ~ factor(SEX), oak_indiv_svy)
model_ols3 <- svyglm(
  log(INCTOT) ~ factor(SEX) +
    AGE + I(AGE ^ 2) +
    relevel(educ_attain, ref ="Highschool diploma"),
  oak_indiv_svy
)
modelsummary(
  list(
    "Earnings gap, no covariates" = model_ols2,
    "Earnings gap, adjusted for age and education" = model_ols3
  ),
  gof_map = gof_stuff,
  exponentiate = TRUE
)
```

Fun!